---
output: github_document
---

<!-- This document is generated from HeightDBH.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Predicting DBH using lidar-derived heights

This report explores the relationship between DBH and height in our field data. Initially, we used the equations from

> Hanus, M.L., D.D. Marshall, and D.W. Hann. 1999. Height-diameter equations for six species in the coastal regions 
of the Pacific Northwest.  Forestry Research Laboratory, contribution 25.  11 p.

to predict DBH given height. This was accomplished by rearranging the equations presented in Hanus et al. to produce a DBH given height
values (derived from the individual trees identified using the lidar data). While this seems to be a reasonable approach, the rearranged
equations introduce bias into the predictions and may not be the best representation of the relationship between height and DBH. We had a 
reviewer for the wind river draft raise this issue but I pretty much ignored the comment since we didn't have a better solution at the time.
I also looked into the equations used in the Pacific Coast variant of FVS as a possible solution. The FVS equations are documented in the 
variant description and are specific to geographic regions within the area covered by the variant. The FVS code provides functions that
give DBH or height but the equations and coefficients are the same. Basically, an equation was fit describing the relationship between
DBH and height (not sure which was used for fitting but I suspect the equation was fit to predict height given DBH) and then rearranged
to solve for the "other" variable.

I developed R functions to do the predictions using both sets of equations along with a pair of functions that allow you to select the method ("hanus"
or "fvs"). You should use this pair of functions as they also handle unit conversions more cleanly than the functions that work with the individual
methods. The code for the prediction functions is in [predictDBH_Height.R](../Rcode/predictDBH_Height.R).

This document is generated using RMarkdown. The source file is [HeightDBH.Rmd](../MiniReports/HeightDBH.Rmd). RMarkdown allows you to mix formatted 
text with R code. It provides a good way to document R code but an even better way to automate report generation that includes R code and inline 
R statements. The reports can include the R code and output or "hide" the R code and just present outputs.

## Load prediction code and field data

We start by loading the code containing the prediction functions and our field data. These are both in the repository but in different folders.
Ally and I measured heights to the highest lidar point in each tree used for the DF/WH species classification. Our method allowed us to consider lean. 
I averaged our height measurements and used this as the
tree height for all further comparisons. Tree height was also computed using an automated method but this method did not consider tree lean. For trees on 
steep slopes, I would expect height errors of 1-3m or more. This error is a combination of tree lean and errors associated with using the ground
elevation under the highest point for the tree (may not be the base of the tree) to compute a height.

```{r}
source("../Rcode/predictDBH_Height.R")

inputDataFile <- "../extras/AdjustedField_T3_Training_TreeTops_AllPlots.csv"

# read data
inputData <- read.csv(inputDataFile, stringsAsFactors = FALSE)

# compute the average height using height measured by Ally and Bob in the point cloud
inputData$aveHt <- (inputData$Total.Height.Ally + inputData$Total.Height.Bob) / 2
```

## Predict DBH for all trees using both methods

I use the Hanus et al. and FVS equations to predict a DBH for all trees.

```{r}
fvsDBH <- c()
hanusDBH <- c()
for (i in 1:nrow(inputData)) {
  fvsDBH[i] <- predictDBH(inputData$Species[i]
                          , inputData$aveHt[i]
                          , method = "fvs"
                          , location = 1
                          , heightUnits = "meters"
                          , DBHUnits = "cm")
  hanusDBH[i] <- predictDBH(inputData$Species[i]
                            , inputData$aveHt[i]
                            , method = "hanus"
                            , location = 1
                            , heightUnits = "meters"
                            , DBHUnits = "cm")
}

inputData$fvsDBH <- fvsDBH
inputData$hanusDBH <- hanusDBH
```

This scatterplot show the relationship between our field-measured DBH and lidar-derived tree heights. Note that these are only trees
where a field tree could be matched with a lidar-derived tree. This means that most (maybe all) trees are in dominant or co-dominant
crown positions. In addition, these data only include Douglas-fir and western hemlock trees. These two species account for about 98% of the 
trees in the field data.

```{r}
DF <- inputData[inputData$Species == "PSME", ]
WH <- inputData[inputData$Species == "TSHE", ]
plot(DF$aveHt, DF$DBH_cm
     , col = "red"
     , pch = 1
     , ylab = "Predicted DBH (cm)"
     , xlab = "Lidar height (m)"
     , main = "Comparing field and predicted DBH & lidar height")
points(WH$aveHt, WH$DBH_cm
       , col = "red", pch = 16)
legend(36, 25, c("Field PSME", "Field TSHE")
       , pch = c(1, 16)
       , col = c("red", "red"))
```

## Compare predictions from the two methods

Next I add the predicted DBH values from the two methods. I think that both methods were fit to predict height given a DBH value so we are
working with rearranged equations in both cases. In general, the Hanus et al. equations produce DBH values that are smaller for a given height
and seem to fit our data better than the FVS equations.

```{r}
plot(DF$aveHt, DF$DBH_cm
     , col = "red", pch = 1
     , ylab = "Predicted DBH (cm)"
     , xlab = "Lidar height (m)"
     , main = "Comparing field and predicted DBH & lidar height")
points(WH$aveHt, WH$DBH_cm, col = "red", pch = 16)
legend(36, 25, c("Field PSME", "Field TSHE")
       , pch = c(1, 16)
       , col = c("red", "red"))

points(DF$aveHt, DF$fvsDBH
       , col = "green", pch = 1)
points(WH$aveHt, WH$fvsDBH
       , col = "green", pch = 16)

points(DF$aveHt, DF$hanusDBH
       , col = "blue", pch = 1)
points(WH$aveHt, WH$hanusDBH
       , col = "blue", pch = 16)

legend(20, 70, c("Hanus PSME", "Hanus TSHE", "FVS PSME", "FVS TSHE")
       , pch = c(1, 16, 1, 16)
       , col = c("blue", "blue", "green", "green"))
```

However, there is some bias in the predictions. Less bias with the Hanus et al. equations compared to the FVS predictions but still 
somewhat concerning. These histograms show the distribution of prediction errors and the titles give the average error (bias).

```{r}
{ # force all lines to execute as one...prevents splitting the code before and after the graphs
par(mfrow=c(2,2))
hist(DF$fvsDBH - DF$DBH_cm
     , main = paste0("FVS DBH error -- DF (bias = ", round(mean(DF$fvsDBH - DF$DBH_cm), 2), " cm)")
     , xlab = "DBH error (cm)")
hist(DF$hanusDBH - DF$DBH_cm
     , main = paste0("Hanus DBH error -- DF (bias = ", round(mean(DF$hanusDBH - DF$DBH_cm), 2), " cm)")
     , xlab = "DBH error (cm)")
hist(WH$fvsDBH - WH$DBH_cm
     , main = paste0("FVS DBH error -- WH (bias = ", round(mean(WH$fvsDBH - WH$DBH_cm), 2), " cm)")
     , xlab = "DBH error (cm)")
hist(WH$hanusDBH - WH$DBH_cm
     , main = paste0("Hanus DBH error -- WH (bias = ", round(mean(WH$hanusDBH - WH$DBH_cm), 2), " cm)")
     , xlab = "DBH error (cm)")
par(mfrow = c(1, 1))
}
```

## Fit new equations

Given that we have field-measured DBH and lidar-derived height values for several hundred trees, it makes sense to try fitting our own equations
to predict DBH given height. First, I tried fitting an equation that predicts height given DBH (this is the same approach used by both FVS and
Hanus et al.). I used the same equation form as that used by Hanus et al. but used height in meters and DBH in cm:

> Height(m) = 1.37 + exp(a~0~ + a~1~(DBH(cm)^a~2~^))

```{r}
DFadd_nls <- nls(data = DF, aveHt ~ 1.37 + exp(a0 + a1 * DBH_cm ^ a2),
                 start = list(a0 = 7.0, a1 = -6.0, a2 = -0.25))
DFnls_coeff <- coef(DFadd_nls)

plot(DF$aveHt, DF$DBH_cm, col = "red", pch = 1, ylab = "Field DBH (cm)", xlab = "Predicted height (m)", main = "Predicted height given field DBH")
points(WH$aveHt, WH$DBH_cm, col = "red", pch = 16)
legend(38, 27, c("Field PSME", "Field TSHE"), pch = c(1, 16), col = c("red", "red"))

# predict heights
DF$predHt <- (log((DF$aveHt - 1.37) / exp(DFnls_coeff["a0"])) / DFnls_coeff["a1"])^(1 / DFnls_coeff["a2"])
                                                                                   
# plot predicted DBH using lidar height
points(DF$aveHt, DF$predHt, col = "black", pch = 1)

# fit new model for WH using Hanus equation form
# DBH in cm and height in m
WHadd_nls <- nls(data = WH, aveHt ~ 1.37 + exp(a0 + a1 * DBH_cm ^ a2),
                 start = list(a0 = 7.0, a1 = -6.0, a2 = -0.25))
WHnls_coeff <- coef(WHadd_nls)

# predict heights
WH$predHt <- (log((WH$aveHt - 1.37) / exp(WHnls_coeff["a0"])) / WHnls_coeff["a1"])^(1 / WHnls_coeff["a2"])

points(WH$aveHt, WH$predHt, col = "black", pch = 16)

# plot predicted height using field DBH
#points(4.5 + exp(WHnls_coeff[1] + WHnls_coeff[2] * WH$DBH_cm ^ WHnls_coeff[3]), WH$DBH_cm, col = "magenta", pch = 16)

legend(38, 36, c("Custom PSME", "Custom TSHE"), pch = c(1, 16), col = c("black", "black"))
```

The resulting predictions seem to do better with our data. However, the predicted heights at the ends of the DBH distribution don't look great. Note
that the graph above has the predicted height on the x axis. this is not conventional but makes comparison with other graphs easier.

Next I fit the same equation form rearranged to predict DBH given height:

> DBH(cm) = log((Height(m) - 1.37) / exp(a~0~)) / a~1~)^(1/a~2~)^

These results do better at the ends of the DBH distribution and appear to fit our data very well.

```{r}
library(nls2)
DFadd_nls <- nls(data = DF, DBH_cm ~ (log((aveHt - 1.37) / exp(a0)) / a1) ^ (1 / a2),
                 start = list(a0 = 7.0, a1 = -6.0, a2 = -0.25))
DFnls_coeff <- coef(DFadd_nls)

plot(DF$aveHt, DF$DBH_cm, col = "red", pch = 1, ylab = "Predicted DBH (cm)", xlab = "Lidar height (m)", main = "Predicted DBH given lidar height")
points(WH$aveHt, WH$DBH_cm, col = "red", pch = 16)
legend(38, 27, c("Field PSME", "Field TSHE"), pch = c(1, 16), col = c("red", "red"))

DF$predDBH <- (log((DF$aveHt - 1.37) / exp(DFnls_coeff["a0"])) / DFnls_coeff["a1"])^(1 / DFnls_coeff["a2"])
points(DF$aveHt, DF$predDBH, col = "orange", pch = 1)

# plot predicted height using field DBH
#points(4.5 + exp(DFnls_coeff[1] + DFnls_coeff[2] * DF$DBH_cm ^ DFnls_coeff[3]), DF$DBH_cm, col = "magenta", pch = 1)

WHadd_nls <- nls(data = WH, DBH_cm ~ (log((aveHt - 1.37) / exp(a0)) / a1) ^ (1 / a2),
                 start = list(a0 = 4, a1 = -28, a2 = -1))
WHnls_coeff <- coef(WHadd_nls)

WH$predDBH <- (log((WH$aveHt - 1.37) / exp(WHnls_coeff["a0"])) / WHnls_coeff["a1"])^(1 / WHnls_coeff["a2"])

points(WH$aveHt, WH$predDBH, col = "orange", pch = 16)

legend(38, 36, c("Custom PSME", "Custom TSHE"), pch = c(1, 16), col = c("orange", "orange"))
```

These histograms show the distribution of prediction errors for the new equation that predicts DBH directly and the titles give the average error (bias).

```{r collapse=TRUE}
{ # force all lines to execute as one...prevents splitting the code before and after the graphs
par(mfrow=c(1,2))
hist(DF$predDBH - DF$DBH_cm
     , main = paste0("DBH error -- DF (bias = ", round(mean(DF$predDBH - DF$DBH_cm), 2), " cm)")
     , xlab = "DBH error (cm)")
hist(WH$predDBH - WH$DBH_cm
     , main = paste0("DBH error -- WH (bias = ", round(mean(WH$predDBH - WH$DBH_cm), 2), " cm)")
     , xlab = "DBH error (cm)")
par(mfrow = c(1, 1))
}
```

## Summary

Overall, fitting new equations using our field data and lidar-derived heights should give us a prediction model useful for the entire T3
study area. The Hanus et al. equations were fit to SMC data from British Columbia, western Washington, and northwestern Oregon. While all of these data
are from the general region, it is well known that conditions on the Olympic Peninsula are different from "typical" westside forest conditions.
The FVS equations were fit using thousands of tree measurements from FIA. The pacific coast variant includes multiple geographic regions with different model
forms and coefficients for each region. While FVS uses a more comprehensive data set, it appears to predict larger DBH values (or shorter heights for 
a given DBH value) than we see in our data.

The final equation to predict DBH directly is of the form:

> DBH(cm) = log((Height(m) - 1.37) / exp(a~0~)) / a~1~)^(1/a~2~)^

Coefficients for DF and WH are provided in the following table.

Species | a~0~ | a~1~ | a~2~
------- | ---- | ---- | ----
DF      | `r round(DFnls_coeff["a0"], 4)` | `r round(DFnls_coeff["a1"], 4)` | `r round(DFnls_coeff["a2"], 4)`
WH      | `r round(WHnls_coeff["a0"], 4)` | `r round(WHnls_coeff["a1"], 4)` | `r round(WHnls_coeff["a2"], 4)`

